---
title: 'Cyclistic: comparing annual members vs. casual riders'
author: "Dario Scardina"
date: "2023-11-24"
output:
  html_document: 
    toc: yes
    css: styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

Blablabla
This is my cap-stone project documentation for Cyclistic case study from the Google Data Analytics Course. 

## Scenario

*Cyclistic* is a bike-sharing company with a fleet of geotracked bycicles locked into a network of different stations in Chicago.
They offer also inclusive bycicles (used by 8% of customers).
The most part of customers ride for leisure, but about 30% use them to commute to work.

They offer 3 different pricing plans:

* single-ride pass (casual rider)
* full-day pass (casual rider)
* annual membership (Cyclistic membership)

## Setup environment

Configuring environment installing and loading necessary libraries:

```{r setup environment}
libraries <- c("tidyverse", "dplyr", "plyr", "readr", "validate", "here", "skimr", "janitor")

missing_libraries <- libraries[!(libraries %in% installed.packages()[,"Package"])]
if(length(missing_libraries) > 0) {
  install.packages(missing_libraries)
}

lapply(libraries, library, character.only = TRUE)
```
Set current working directory:
```{r set working directory}
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path ))
```

## Load dataset
Set number of files to load (1 = just load January 2022, 12 = complete analysis of 2022).
Note that files are sorted by name, so if num_files = 3, it will load first trimester of 2022.
Using lower num_files speed-up analysis.
```{r choose number of files to load}
num_files_to_load = 1
```

Load csv files and merge the dataset in a single dataframe:
```{r load and merge files}
df <- list.files(path = "dataset",  # Identify all CSV files in relative folder
                       pattern = "*.csv", full.names = TRUE) %>% 
  sort() %>%
  head(num_files_to_load) %>%
  lapply(read_csv) %>% # Store all files in list: apply read_csv to each csv and combine them into a list
  imap(~ .x %>% mutate(file_title = .y)) %>% # add column with the index of file TODO enhance with the name of the file
  bind_rows                                         # Combine data sets into one data set: bind any number of data frames by row
```
## How do annual members and casual riders use Cyclistic bikes differently?
In order to answer the question, it is necessary to produce a report with the following deliverables:

1. A clear statement of business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. An analysis summary
5. Supporting visualizations and key findings
6. Top three reccomendations bases on the analysis

## Ask phase
In this phase the first step consist to identify the problem we're trying to solve.
*Cyclistic* is a bike-sharing company with a fleet of geotracked bycicles locked into a network of different stations in Chicago.
They offer also inclusive bycicles (used by 8% of customers).
The most part of customers ride for leisure, but about 30% use them to commute to work.

They offer 3 different pricing plans:
The goal of this section is to define a clear statement of business task.

* single-ride pass (casual rider)
* full-day pass (casual rider)
* annual membership (Cyclistic membership)
The first step consist to identify the problem we're trying to solve, starting 
from considerations made by our 'imaginary' colleagues.

In order to garantuee the future growth, Company's stakeholders think that a marketing 
campaign to convert casual riders to Cyclistyc Members would be the key of success.

This hypotesis? is based on the following assumption:
This hypotesis is based on the following assumption:

* Cyclistic’s finance analysts have concluded that **annual members are much more profitable than casual riders**
* **Converting casual riders into annual members** will lead to growth
* Cyclistic’s finance analysts have concluded that *annual members are much more profitable than casual riders*
* *Converting casual riders into annual members* will lead to growth

In our case we have to answer the following question: How do annual members and casual riders use Cyclistic bikes differently?
In our case we have to answer the following question: *How do annual members and casual riders use Cyclistic bikes differently?*
The problem we are trying to solve is to identify is understand if there is a different use of the bikes between members and casual riders in terms of how, when and why they use the service.
Knowing that could be useful to understand why casual riders prefer to use single or daily pass instead of buying membership.

CLEAR STATEMENT OF BUSINESS TASK

## Prepare

### Guiding questions

#### Where the data is located?

You will use Cyclistic’s historical trip data to analyze and identify trends. Download the Cyclistic trip data here. (Note: The datasets
have a di erent name because Cyclistic is a  ctional company. For the purposes of this case study, the datasets are appropriate
and will enable you to answer the business questions).

#### How data is organized

In the website there is a list of zip files. Each of the zip refers to:

* tripdata cos'è? Useremo l'anno "completo" più recente. 2022
* Stations: cartella compressa che contiene 2 csv uno con i trips e uno con le stations. stations include identificativo con coordinate geografiche e altre info.
* Trips: sembrano avere la stessa struttura di trip data. Contengono i dati per quarter.

Each file seems to refer to a specific period (fare verifica su singolo file) that could be (a seconda del caso):

* YEAR-MONTH
* YEAR-QUARTER
* YEAR-SEMESTER: union of two consecutive quarters
* YEAR

Abbiamo data last modified e dimensione del file.

#### Are there issues with bias or credibility in this data? Does your data ROCCC?

Data is **O**riginal: it is provided by Motivate International Inc. Lyft Bikes and Scooters, LLC (“Bikeshare”) that operates the City of Chicago’s (“City”) Divvy bicycle sharing service through its official website.
Data is **C**omprehensive: it contains all the information necessary for this analysis. In particular, as reported in the source webpage https://divvybikes-marketing-staging.lyft.net/system-data, for a brief recap:
Each trip is anonymized and includes:

* Trip start day and time
* Trip end day and time
* Trip start station
* Trip end station
* Rider type (Member, Single Ride, and Day Pass)

Data is **C**urrent: dataset is constantly updated every month with last data produced.

Data is **C**ited: it is provided by the bike-sharing company that operates its business on Chicago (U.S.).


#### How are you addressing licensing, privacy, security, and accessibility?

The data has been made available by Motivate International Inc. under this license. vedi pdf.
The data for 2022 does not contains any sensible or personal information but only info about trips.
See info https://divvybikes-marketing-staging.lyft.net/system-data

#### How did you verify the data’s integrity?

#### How does it help you answer your question?

#### Are there any problems with the data?

* Molti file da scaricare: come automatizzare?
* Alcuni csv troppo pesanti (non gestibili da Gsheet)
* csv del 2022 alcuni hanno "" altri no. Capire come R li gestisce.

### Key tasks

1. Download and store data appropriately
2. Identify how it's organized
3. Sort and filter the data
4. Determine the credibility of the data

### Deliverable 

A description of all data sources used

## Process
 
### Guiding questions

* What tools are you choosing and why? Gsheet cannot handle all the csv files because some are too big (even if loaded, managing is really hard: waiting time ...)
* Have you ensured your data’s integrity? 
* What steps have you taken to ensure that your data is clean? 

First check when reading csv:

* chr  (7): ride_id, rideable_type, start_station_name, start_station_id, end_station_name, end_station_id, member_casual
* dbl  (4): start_lat, start_lng, end_lat, end_lng
* dttm (2): started_at, ended_at

1. ride_id: must be a guid of 16 characters.
2. rideable_type: 3 values electric, classic or docked bike.
3. started_at and ended_at: timestamps
4. start_station_ame and end_station_name: strings
5. start_station_id and end_station_id: id not clear (to investigate: sometimes is only 3 digits, sometimes starts with 2 letters)
6. _lng fields: valid numeric format for latitude and longitude (from -90 to 90)

* How can you verify that your data is clean and ready to analyze?

Innanzi tutto partiamo dal seguente comando che ci da visione d'insieme del dataset:
```{r}
skim(df)
```
Per il gruppo di colonne character, la prima cosa che salta all'occhio è che il complete rate non è 1 (quindi abbiamo dei missing values).

Consideriamo intanto la start_station.

Eseguiamo il comando

missing_start_station_geolocation <- df %>%  filter(is.na(start_station_name)) %>% select(start_lat, start_lng)

per filtrare tutte le righe 

* Have you documented your cleaning process so you can review and share those results?

### Key tasks
1. Check the data for errors.
2. Choose your tools.
3. Transform the data so you can work with it effectively.
4. Document the cleaning process.

### Deliverable
Documentation of any cleaning or manipulation of data

## Analyze

### Guiding questions

* How should you organize your data to perform analysis on it?
* Has your data been properly formaed?
* What surprises did you discover in the data?
* What trends or relationships did you nd in the data?
* How will these insights help answer your business questions?

### Key tasks
1. Aggregate your data so it’s useful and accessible.
2. Organize and format your data.
3. Perform calculations.
4. Identify trends and relationships.

### Deriverable
A summary of the analysis.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```


